A free/open-source Kazakh-Tatar machine translation system

### 2013-04-22 paper/paper.tex
error analysis
new stuff cleaning blah

\newcommand{\todo}[1]{\{\textbf{TODO: #1}\}}

@@

\begin{figure}
\begin{lstlisting}
<rule> 
  <match lemma="ақпарат" tags="n.attr"/> 
  <match lemma="құрал" tags="n.*.px3sp.*">
    <select lemma="чара"/>
  </match> 
</rule>
\end{lstlisting}

\todo{A couple more lexical-selection rules }
\caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
\label{fig:lrx}
%\vspace{-1em}
\end{figure}

@@

The system currently has a total of 33 lexical selction rules.

@@

Some of the corrections that were made as part of postediting were were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.  A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.

\begin{figure}[htbp]
	\begin{tabular}{lll}
		\toprule
		form & current & expected \\\midrule
		ди\tag{<v>}\tag{<tv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & дия & ди \\
		йөр\tag{<v>}\tag{<iv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & йөрә & йөри \\
		кияү\tag{<n>}\tag{<px3sp>}\tag{<nom>} & кияүе & кияве \\
		укы\tag{<v>}\tag{<iv>}\tag{<ger>}\tag{<nom>} & укыу & уку \\\bottomrule
	\end{tabular}
	\caption{Examples of some phonological problems in the Tatar transducer.}
	\label{tab:phonerrors}
\end{figure}

\todo{Example of lack of phon rule}

@@

\begin{table}
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
  \midrule
  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
  \bottomrule
  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
  \end{tabular}
    \caption{Word error rate over two corpora. The \texttt{devel} `development' corpus was used during development 
       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.}
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}

@@

%The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
%\begin{itemise}
%  \item Combinations of case and possessive suffixes,
%  \item Orthographical representations of phonology,
%  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
%\end{itemise}

%\cite{TyersAlperen2010}

>>>>

%\newcommand{\todo}[1]{\{\textbf{TODO: #1}\}}
\newcommand{\todo}[1]{}

@@

\begin{figure}
%	\subfloat[A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.]{
\begin{lstlisting}
<rule> 
  <match lemma="ақпарат" tags="n.attr"/> 
  <match lemma="құрал" tags="n.*.px3sp.*">
    <select lemma="чара"/>
  </match> 
</rule>
\end{lstlisting}
%}\\
%\subfloat[]{
%}
\todo{A couple more lexical-selection rules }
%\caption{Some lexical selection rules in Kazakh}
\caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
\label{fig:lrx}
%\vspace{-1em}
\end{figure}

@@

Another lexical selection rule is shown in figure~\ref{fig:lrx2}.  This rule ensures that the Kazakh word рет \eng{row, time(s)} is translated as тапкыр \eng{time(s)} when it is used with a number; otherwise it translates as рәт \eng{row}.
\begin{figure}
\begin{lstlisting}
  <rule>
    <match tags="num"/>
    <match lemma="рет" tags="n.*">
        <select lemma="тапкыр"/>
    </match>
  </rule>
\end{lstlisting}
\caption{A lexical selection rule that selects Tatar тапкыр for Kazakh рет when preceded by a number.}
\label{fig:lrx2}
\end{figure}

Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).  The lexical selection rule is provided in figure~\ref{fig:lrx3} chooses the latter Tatar translation after the word рухани \eng{soul}.

\begin{figure}
\begin{lstlisting}
  <rule>
    <match lemma="рухани" tags="adj"/>
    <match lemma="өмір" tags="n.*">
        <select lemma="тормыш"/>
    </match>
  </rule>
\end{lstlisting}
\caption{A lexical selection rule that selects Tatar тормыш for Kazakh өмір when preceded by the word рухани.}
\label{fig:lrx3}
\end{figure}

The system currently has a total of 33 lexical selection rules.

@@

Some of the corrections that were made as part of postediting were were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.  A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.  Aside from the irregular behaviour of йөр, which would need its irregularities implemented through special phonotactics in lexc, these issues are all shortcomings of our phonological layer, implemented in twol.

\begin{table}[htbp]
	\centering
	\begin{tabular}{lll}
		\toprule
		form & current & expected \\\midrule
		ди\tag{<v>}\tag{<tv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & дия & ди \\
		йөр\tag{<v>}\tag{<iv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & йөрә & йөри \\
		кияү\tag{<n>}\tag{<px3sp>}\tag{<nom>} & кияүе & кияве \\
		укы\tag{<v>}\tag{<iv>}\tag{<ger>}\tag{<nom>} & укыу & уку \\\bottomrule
	\end{tabular}
	\caption{Examples of some phonological problems in the Tatar transducer.}
	\label{tab:phonerrors}
\end{table}

@@

\begin{table}
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
  \midrule
  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over two corpora. The development corpus (\texttt{devel}) was used during development 
       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.}
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}

@@

%The majority of errors currently are due to one of the following:
%\begin{itemise}
%   \item mistakes and gaps in the Tatar morphophonology component,
%   \item lack of transfer rules to hande some of the Kazakh compund tense forms, and
%   \item disambiguation errors.
%%  \item Combinations of case and possessive suffixes,
%%  \item Orthographical representations of phonology,
%%  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
%\end{itemise}

The majority of errors currently are due to one of the following: mistakes and gaps in the Tatar morphophonology component, lack of transfer rules to handle some Kazakh compounds, and disambiguation errors.
%\cite{TyersAlperen2010}
